Type-Safe Modular Parsing

### 2017-09-04 SLE_17/paper.tex
Working on final version.

%% For double-blind review submission
%% \documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=false}
%% For single-blind review submission
\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

@@

\settopmatter{printacmref=false}

@@

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

@@

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

%\ccsdesc[500]{Software and its engineering~General programming languages}
%\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
%\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional

@@

>>>>

%% For double-blind review submission
%% \documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=false}
%% For single-blind review submission
\documentclass[sigplan,screen]{acmart}\settopmatter{}
%% For final camera-ready submission
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

@@

@@

%%% The following is specific to SLE'17 and the paper
%%% 'Type-Safe Modular Parsing'
%%% by Haoyuan Zhang, Huang Li, and Bruno C. d. S. Oliveira.
%%%
%\setcopyright{ACMCT}
\setcopyright{acmcopyright}
\acmPrice{15.00}
\acmDOI{10.1145/3136014.3136016}
\acmYear{2017}
\copyrightyear{2017}
\acmISBN{978-1-4503-5525-4/17/10}
\acmConference[SLE'17]{2017 ACM SIGPLAN International Conference on Software Language Engineering}{October 23--24, 2017}{Vancouver, Canada}

@@

\begin{CCSXML}
	<ccs2012>
	<concept>
	<concept_id>10011007.10011006.10011008.10011009.10011011</concept_id>
	<concept_desc>Software and its engineering~Object oriented languages</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	<concept>
	<concept_id>10003752.10010124.10010125.10010128</concept_id>
	<concept_desc>Theory of computation~Object oriented constructs</concept_desc>
	<concept_significance>300</concept_significance>
	</concept>
	</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Object oriented languages}
\ccsdesc[300]{Theory of computation~Object oriented constructs}

%% Keywords
%% comma separated list
\keywords{modular parsing, object algebras, semantic modularity}  %% \keywords is optional

@@

\section*{Acknowledgements}
We would like to thank the anonymous reviewers for their helpful comments. This work has been sponsored by the Hong Kong Research Grant Council projects number 27200514 and 17258816.

### 2017-09-04 SLE_17/sections/Introduction.tex
Working on final version.

To address the modularity limitations of Programming Languages, several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use.
Examples include many tools for developing Feature-Oriented Software-Product
Lines (SPLs)~\cite{AK:JOT09,Kastner11road}, some Language Workbenches~\cite{Erdweg201524}, or extensible parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{AK:JOT09} to
enable the development modular program features.
%Such
%composition techniques collect the code for multiple features and
%merge them together when a concrete combination of features is needed
%for a particular program.
As Kastner et. al~\cite{Kastner11road} note,
a typical drawback of feature-oriented SPL implementations, which
more generally applies to syntactic modularity approaches, is that
such ``\emph{implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''.

>>>>

To address the modularity limitations of Programming Languages, several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use.
Examples include many tools for developing Feature-Oriented Software-Product
Lines (SPLs)~\cite{AK:JOT09,Kastner11road}, some Language Workbenches~\cite{Erdweg201524}, or extensible parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Viera2012,Warth2016,schwerdfeger09}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{AK:JOT09} to
enable the development modular program features.
%Such
%composition techniques collect the code for multiple features and
%merge them together when a concrete combination of features is needed
%for a particular program.
As Kastner et. al~\cite{Kastner11road} note,
a typical drawback of feature-oriented SPL implementations, which
more generally applies to syntactic modularity approaches, is that
such ``\emph{implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''.

### 2017-09-04 SLE_17/sections/RelatedWork.tex
Working on final version.

\paragraph{Syntactically Extensible Parsing}
Extensible parser generators~\cite{antlr1995,Grimm2006,Gouseti2014,Viera2012,Warth2016}
are a mainstream area of modular syntax and parsing. They allow users to write
modular grammars, where new
non-terminals and production rules can be introduced, some can even
override existing rules in the old grammar modules. For instance,
\textit{Rats!}~\cite{Grimm2006} constructs its own module system for
the collection of grammars, while NOA~\cite{Gouseti2014} uses Java
annotation to collect all information before producing an ANTLR~\cite{antlr1995} grammar
and the parsing code. Those parser
generators focus on the \textit{syntactic extensibility} of grammars:
they rely on whole compilation to generate a global parser, even if
there is only a slight modification in the grammar. Some of those
parser generators may statically check the correctness and unambiguity
of grammars. In contrast, because our approach is based on parser
combinators, there is no support for ambiguity checking.  However, as
far as we are aware, no extensible parser generators support separate
compilation or modular type-checking. It is worth mentioning that in~\cite{Viera2012}, users can define
grammar fragments as typed Haskell values, and combine them on the fly. Later they are processed
by a typed parser generator. Nevertheless this requires
a lot of advanced language features, making client complex. Our
approach is simple and a straightforward use of OO programming, 
and makes parsing code directly reusable.

@@

@@

\paragraph{Extensible Parsing Algorithms}
\textit{Parse table composition}~\cite{bravenboer2008parse}
is an approach where grammars are compiled to
modular parse tables. Those parse tables are expressed as DFAs
or NFAs, and later they can be composed by an algorithm, to provide
separate compilation for parsing. The generation of parse tables can
be quite expensive in terms of performance. The approach
is quite different from ours, since it uses parse
 tables, whereas we use parser combinators.
Our approach supports both
\emph{separate compilation} as well as \emph{modular
  type-checking}. Moreover, the extensibility of parsing is further
available at language composition and lexical level.

>>>>

\paragraph{Syntactically Extensible Parsing}
Extensible parser generators~\cite{antlr1995,Grimm2006,Gouseti2014,Viera2012,Warth2016,schwerdfeger09}
are a mainstream area of modular syntax and parsing. They allow users to write
modular grammars, where new
non-terminals and production rules can be introduced, some can even
override existing rules in the old grammar modules. For instance,
\textit{Rats!}~\cite{Grimm2006} constructs its own module system for
the collection of grammars, while NOA~\cite{Gouseti2014} uses Java
annotation to collect all information before producing an ANTLR~\cite{antlr1995} grammar
and the parsing code. Those parser
generators focus on the \textit{syntactic extensibility} of grammars:
they rely on whole compilation to generate a global parser, even if
there is only a slight modification in the grammar. Some of those
parser generators may statically check the correctness and unambiguity
of grammars. In contrast, because our approach is based on parser
combinators, there is no support for ambiguity checking.  However, as
far as we are aware, no extensible parser generators support separate
compilation or modular type-checking. It is worth mentioning that in~\cite{Viera2012}, users can define
grammar fragments as typed Haskell values, and combine them on the fly. Later they are processed
by a typed parser generator. Nevertheless this requires
a lot of advanced language features, making client complex. Our
approach is simple and a straightforward use of OO programming, 
and makes parsing code directly reusable.

@@

Besides, type-specific languages~\cite{omar14} do support
separate compilation and type safety. The modularity of parsing is incorporated in its type
system, which allows users to introduce new syntax for parsing. Whereas in our approach, parsing 
strategy is an orthogonal issue since we rely on an external parsing library.
Our approach does not rely on an elaborated semantics extension, but simply uses a pattern
that is applicable in mainstream OO languages, where the type system helps ensure
type safety.

@@

\paragraph{Extensible Parsing Algorithms}
\textit{Parse table composition}~\cite{bravenboer2008parse, schwerdfeger2010}
is an approach where grammars are compiled to
modular parse tables. Those parse tables are expressed as DFAs
or NFAs, and later they can be composed by an algorithm, to provide
separate compilation for parsing. The generation of parse tables can
be quite expensive in terms of performance. The approach
is quite different from ours, since it uses parse
 tables, whereas we use parser combinators.
Our approach supports both
\emph{separate compilation} as well as \emph{modular
  type-checking}, and is applied in common OO languages. Moreover, the extensibility of parsing is further
available at language composition.
